Flat-histogram method comparison on the 2D Ising model

### 2020-06-18 ising/referee-response.md

Dear editor,

We appreciate the referees' careful reading of our paper, and feel that their
feedback has led to a significant improvement. We have addressed all of the
first referee's comments by added two paragraphs to the introduction and a
section in the paper to 'Address the general problem of all "flat-histogram"
methods'. We also strengthen the abstract to address the concerns raised by the
second referee. We point out that: (i) Our paper is the first of its kind (to
the best of our knowledge) to compare ``production run'' WL with other
flat-histogram methods. (ii) It is important to test SAD's convergence
properties on the 2D Ising where the range of energies is easily known (thus
biasing in favor of WL methods). ...

Response to first referee
# -----------------------------------------------------------------------------#

The first referee points out that 'Simulation set-up and analysis protocols are
sound and give confidence in the quality of the presented data and
conclusions.'

The referee also asks that three additions be made to the paper.

(i) It would be 'worthwhile to work out the [slower convergence for SAD due to
the energy range being an a priori parameter in WL/MUCA] in the paper, and, if
possible, quantify the loss in efficiency coming from energy range
determination.'

We have added a paragraph? subsection in the SAD/results portion of the paper that
address this.

(ii) Address the general problem of all "flat histogram" methods that the
performance of these methods is due to hidden barriers almost always lower than
what one would expect from a random walk in energy (in most cases still
exponentially growing). This has been discussed by Nadler et al (PRE 75 (2007)
026109) who also discus a protocol for optimizing the ensemble which minimizes
the residual exponential slowing down. The authors should point out if and how
SAD can be adapted to such "non-flat-histogram" methods.

We have added a paragraph that 'Address the general problem of all "flat
histogram" methods' in the introduction.

(iii) Add a few short remarks on comparison with replica exchange methods would
be helpful.

We have added a 'few short remarks' on comparison with replica exchange in the
introduction.

Response to second referee
# -----------------------------------------------------------------------------#

We thank the second referee for bringing to our attention two points so that we
can better address them in our paper and make them completely clear.

(i) The work is almost a repetition of what the authors did when they
introduced the algorithm.

Our paper is the first of its kind (to the best of our knowledge) to compare
``production run'' WL with other flat-histogram methods. The first (not the second) referee
notes rather strongly that this is the only way that WL should be done;
however, no research work has ever compared this with other flat-histogram
methods and certainly not with the completeness shown here. Also, the very fact that
it is compared with pure WL is proof for the first referee's statement that
"this is how WL should be done" since it performs superior to pure WL.

(ii) It is not clear[ly] understood why the current calculations were not
included in the original article. Although the results presented by the
[authors] are original, they are only a simple extension of the first work, and
do not meet the standards of PRE readers.

The second referee an important question here to which there are two answers.
First, we only recently became aware of ``production run'' WL and wanted to
rapidly compare this with other histogram methods due to the lack of literature
detailing its performance vs other flat-histogram methods. Second, SAD was
initially tested on methods where it could benefit from knowing the temperature
range of interest in advance. It is critical to test it's convergence
properties on a different type of system where the range of energies is easily
known (thus biasing in favor of WL methods).

In response to the second referee's comments, we have better highlighted these
important points in the abstract.

### 2020-06-18 ising/referee-response.md

The referee also asks that three additions be made to the paper.

(i) It would be 'worthwhile to work out the [slower convergence for SAD due to
the energy range being an a priori parameter in WL/MUCA] in the paper, and, if
possible, quantify the loss in efficiency coming from energy range
determination.'

We have added a paragraph? subsection in the SAD/results portion of the paper that
address this.

## Ideas

- Measure time until we last find a new important energy
- Find the fraction of moves (in the end) that are outside the range of interesting energies.
- Discuss in results or discussion

(ii) Address the general problem of all "flat histogram" methods that the
performance of these methods is due to hidden barriers almost always lower than
what one would expect from a random walk in energy (in most cases still
exponentially growing). This has been discussed by Nadler et al (PRE 75 (2007)
1)      who also discus a protocol for optimizing the ensemble which minimizes
the residual exponential slowing down. The authors should point out if and how
SAD can be adapted to such "non-flat-histogram" methods.

@@

We have added a paragraph that 'Address the general problem of all "flat
histogram" methods' in the introduction.

## Ideas

- Read this paper (probably both of us) and write it up.  We need to describe the paper and discuss how this relates.
- My guess is that SAD is like any of the other methods for non-flat-histogram approaches.
- (Optimized ensemble *follows* a flat histogram method.)

@@

(iii) Add a few short remarks on comparison with replica exchange methods would
be helpful.

We have added a 'few short remarks' on comparison with replica exchange in the
introduction.

## This is a bit more work

- Our method *should* be amenable to replica exchange parallelization, just as WL is.

@@

We thank the second referee for bringing to our attention two points so that we
can better address them in our paper and make them completely clear.

(i) The work is almost a repetition of what the authors did when they
introduced the algorithm.

Our paper is the first of its kind (to the best of our knowledge) to compare
``production run'' WL with other flat-histogram methods. The first (not the second) referee
notes rather strongly that this is the only way that WL should be done;
however, no research work has ever compared this with other flat-histogram
methods and certainly not with the completeness shown here. Also, the very fact that
it is compared with pure WL is proof for the first referee's statement that
"this is how WL should be done" since it performs superior to pure WL.

## Notes

- We think it is important to bridge the disconnect between *users* of WL and developers of improved-flat histogram methods.
- Maybe two paragraphs? THIS IS REALLY IMPORTANT (IN THE LETTER)

@@

(ii) It is not clear[ly] understood why the current calculations were not
included in the original article. Although the results presented by the
[authors] are original, they are only a simple extension of the first work, and
do not meet the standards of PRE readers.

The second referee asks an important question here to which there are two answers.
First, we only recently became aware of ``production run'' WL and wanted to
rapidly compare this with other histogram methods due to the lack of literature
detailing its performance vs other flat-histogram methods. Second, SAD was
initially tested on methods where it could benefit from knowing the temperature
range of interest in advance. It is critical to test it's convergence
properties on a different type of system where the range of energies is easily
known (thus biasing in favor of WL methods).

## Notes

- Two paragraphs for two answers?
- We only became aware of "production run" WL as a result of the response to our
  first paper.  As mentioned (hinted at) in the acknowledgement, XXX emailed us after
  reading our paper and suggested that we test ...
- 2nd: Ising is different.  Range of energies is known.  We omitted it in ther first
  paper due to size and time.,  These computations are lengthy
  and slow.

### 2020-06-20 ising/referee-response.md

Response to first referee
# -----------------------------------------------------------------------------#

The first referee points out that 'Simulation set-up and analysis protocols are
sound and give confidence in the quality of the presented data and
conclusions.'

The referee also asks that three additions be made to the paper.

(i) "It would be worthwhile to work out the [slower convergence for SAD due to
the energy range being an a priori parameter in WL/MUCA] in the paper, and, if
possible, quantify the loss in efficiency coming from energy range
determination."

We have added a paragraph in the discussion section that explains the loss in
efficiency SAD suffers by not specifying an energy range a priori. We are able
to quantify this by identifying the last time SAD finds a new important energy
and then finding the fraction of moves (in the end) that are outside the range
of interesting energies.

(ii) "Address the general problem of all "flat histogram" methods that the
performance of these methods is due to hidden barriers almost always lower than
what one would expect from a random walk in energy (in most cases still
exponentially growing). This has been discussed by Nadler et al (PRE 75 (2007)
1) who also discus a protocol for optimizing the ensemble which minimizes the
residual exponential slowing down. The authors should point out if and how SAD
can be adapted to such "non-flat-histogram" methods."

We have added a paragraph that 'Address the general problem of all "flat
histogram" methods' in the introduction. We discuss how the performance is
impacted due to hidden barriers. We also add that WL has already been adapted
to replica-exchange and point out that this should be possible for SAD as well.

@@

(iii) Add a few short remarks on comparison with replica exchange methods would
be helpful.

We have added a 'few short remarks' on comparison with replica exchange in the
introduction. WL has been formulated as replica-exchange WLMC whereby it adopts
the concept of conformational swapping directly from parallel tempering. Each replica
is allowed to traverse the entire energy range via overlapping energy windows.

While we have performed limited comparison with replica exchange methods (in
our previous work we found the heat capacity to be consistent with computations
performed with replica exchange), SAD could be extended in much the same way
that WL was. In fact, this would make an interesting comparison with replica
exchange statistical temperature Monte Carlo (as they both specify a
temperature range) and should be studied on a temperature range system in a
future work.

@@

Response to second referee
# -----------------------------------------------------------------------------#

We thank the second referee for bringing to our attention two points so that we
can better address them in our paper and make them completely clear.

(i) "The work is almost a repetition of what the authors did when they
introduced the algorithm."

@@

Our paper is the first of its kind (to the best of our knowledge) to compare
``production run'' WL with other flat-histogram methods. The first (not the
second) referee notes rather strongly that this is the only way that WL should
be done; however, no research work has ever compared this with other
flat-histogram methods and certainly not with the completeness shown here.
Also, the very fact that it is compared with pure WL is proof for the first
referee's statement that "this is how WL should be done" since it performs
superior to pure WL.

We believe that this paper helps by bridging the disconnect between *users* of
WL and developers of improved flat-histogram methods. The majority of
developers use pure WL when comparing multiple flat-histogram methods since
there is no direct comparison of WL followed by a production run with methods
such as 1/t-WL or SAMC.

(ii) "It is not clear[ly] understood why the current calculations were not
included in the original article. Although the results presented by the
[authors] are original, they are only a simple extension of the first work, and
do not meet the standards of PRE readers."

The second referee asks an important question here to which there are two
answers. We only became aware of "production run" WL as a result of the
response to our first paper. As mentioned (hinted at) in the acknowledgement,
Johannes Zierenberg emailed us after reading our paper and suggested that we
test WL followed by a production run with a number of flat-histogram methods.
Due to the lack of literature detailing its performance vs other flat-histogram
methods, the 2D Ising model seems the logical starting point for such a
comparison.

Second, SAD was initially tested on methods where it could benefit from knowing
the temperature range of interest in advance. It is critical to test it's
convergence properties on a different type of system where the range of
energies is easily known (thus biasing in favor of WL methods). Also in the
case of the 128 X 128 2D Ising system, the computation time is quite lengthy
(more than 6 months for some calculations) representing a superior and robust
comparison result.

In response to the second referee's comments, we have better highlighted these
important points in the abstract and conclusion.

### 2020-06-20 ising/ising-body.tex
made some of the changes. Needs some work.

@@

\section{Conclusion}
We find that SAD, $1/t$-WL, and WL followed by a production run (with an
adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure
WL do not suffer from error saturation. We find that for larger Ising
systems SAD reduces the update factor more slowly (and conservatively)
than $1/t$-WL and WL followed by a production run. This means that SAD
will take proportionately more moves to converge to the same value as
$1/t$-WL as system size is increased.  However, the WL methods are
solving a simpler problem because they are given the energy range
\emph{a priori}, rather than a temperature range of interest such as
SAD requires.  The 2D Ising model is a system where the energy range
is known \emph{a priori.}, which makes a comparison between WL methods
and SAD more biased in favor of WL methods. For the general case in
which the energy range of the system is not known and where a range of
desired temperatures is known, this will make the SAD method
considerably more convenient, and quite possibly more efficient than a
process involving multiple simulations to determine the energy range
of interest. This work also dramatically demonstrates that WL followed
by a production run performs extremely well and is preferable to pure
WL for ensuring both ergodicity and detailed balance.

>>>>

Nadler \emph{et al.} and Trebst \emph{et al.} systematically examined optimized
ensembles due to performance issues flat-histogram methods faced when
confronted with hidden energy barriers~\cite{nadler2007generalized,
nadler2007dynamics, trebst2004optimizing}. Trebst \emph{et al.} used their
optimized ensemble to systematically set the temperature range(s) and
implemented an optimized replica-exchange~\cite{trebst2006optimized}.
Replica-exchange Wang-Landau has been developed with great
success~\cite{perera2016magnetic} and adapting flat-histogram methods to use
such an optimized ensemble is doable

Last time for 32 X 32 is TL 2047388. Do I look at histogram to figure this out???

But optimized ensemble requires the energy range???

@@

\section{Conclusion}
We find that SAD, $1/t$-WL, and WL followed by a production run (with an
adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure
WL do not suffer from error saturation. We find that for larger Ising
systems SAD reduces the update factor more slowly (and conservatively)
than $1/t$-WL and WL followed by a production run. This means that SAD
will take proportionately more moves to converge to the same value as
$1/t$-WL as system size is increased.  However, the WL methods are
solving a simpler problem because they are given the energy range
\emph{a priori}, rather than a temperature range of interest such as
SAD requires.  The 2D Ising model is a system where the energy range
is known \emph{a priori.}, which makes a comparison between WL methods
and SAD more biased in favor of WL methods. For the general case in
which the energy range of the system is not known and where a range of
desired temperatures is known, this will make the SAD method
considerably more convenient, and quite possibly more efficient than a
process involving multiple simulations to determine the energy range
of interest. This work also dramatically demonstrates that WL followed
by a production run performs extremely well and is preferable to pure
WL for ensuring both ergodicity and detailed balance. While this has long
been thought to be the case and that a multi-canonical run where WL is used to
determine the parameters is the only way WL should be used, this research
represents the first conclusive comparison among all of these
flat-histogram methods.

### 2020-06-20 ising/ising.tex
fix ising paper so that it updates from ising-body.tex.
made some of the changes. Needs some work.

\begin{abstract}
  We compare the convergence of several flat-histogram methods applied
  to the 2D Ising model, including the recently introduced stochastic
  approximation with a dynamic update factor (SAD) method.  We compare
  this method with the Wang-Landau (WL) method, the $1/t$ variant of
  the WL method, and standard stochastic approximation Monte Carlo
  (SAMC).  In addition, we consider a procedure WL followed by a
  ``production run'' with fixed weights that refines the estimation of
  the entropy.  To our knowledge, this work is the first to test this
  approach against other methods.
  %% To our knowledge, the
  %% convergence rate of this production-run WL approach has never been 
%% We examine the convergence of stochastic approximation with a dynamic
%% update factor (SAD) with other flat-histogram methods applied to the 2D
%% Ising model. The comparison with stochastic approximation Monte Carlo (SAMC)
%% and Wang-Landau (WL) methods shows that SAD performs robustly and without user
%% input knowledge of an energy range. We also compare SAD with WL followed by a
%% ``production run'' with fixed weights which is the first comparison of its kind
%% recorded.
We find that WL followed by a production run \emph{does} converge to
the true density of states, in contrast to pure WL.  Three of the
methods converge robustly: SAD, $1/t$-WL, and WL followed by a
production run.  Of these, SAD does not require \emph{a priori}
knowledge of the energy range.
\end{abstract}

>>>>

\begin{abstract}

We compare the convergence of several flat-histogram methods applied to the 2D
Ising model, including the recently introduced stochastic approximation with a
dynamic update factor (SAD) method. We compare this method with the Wang-Landau
(WL) method, the $1/t$ variant of the WL method, and standard stochastic
approximation Monte Carlo (SAMC). In addition, we consider a procedure WL
followed by a ``production run'' with fixed weights that refines the estimation
of the entropy. To our knowledge, this work is the first to test this approach
against other methods. We find that WL followed by a production run \emph{does}
converge to the true density of states, in contrast to pure WL. Three of the
methods converge robustly: SAD, $1/t$-WL, and WL followed by a production run.
Of these, SAD does not require \emph{a priori} knowledge of the energy range.
This work also shows that WL followed by a production run performs superior to
other forms of WL while ensuring both ergodicity and detailed balance.

\end{abstract}
